---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Yoon, Justin"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

str(mydata)

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS, mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: By just looking at the summary of the data, the WHOLE, SHUCK, and VOLUME variables have mins that are significantly lower than the mean/median and maxs that significantly higher than the mean/median. This may contribute to potential skewness and outliers***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
addmargins(table(mydata$SEX,mydata$CLASS))
barplot(height = table(mydata$SEX,mydata$CLASS),legend.text = c("Female","Infant","Male"), main = "CLASS membership,SEX-differentiated",ylab = "Frequency", col = c("red","blue","green"), beside = TRUE)
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: Based on the distribution of the plot, the majority of the infants reside in A2 with a medium to high frequency in A1 and A3. The distribution of the infants skews to the right. The males and females have the opposite distribution. They have a majority residing in A3 with a medium frequency in A4 and A5. The distribution of males and females skews to the left but is slightly more uniform. A3 has the greatest frequency of abalones with the highest frequency of males and females and a medium frequency of infants as they start to decline. The sex distribution of abalones by class makes sense based on the age classes***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
work <- mydata[sample(x = 1:nrow(mydata), size = 200, replace = FALSE),]
plot(work[,2:6])
```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
plot(x = mydata$VOLUME, y = mydata$WHOLE, col = c("blue","green","red","orange", "grey")[mydata$CLASS], main = "Whole weight, in grams|Volume, cm^3", xlab = "Volume",ylab = "Whole weight", pch = 20)
legend("bottomright", c("A1","A2","A3","A4", "A5"), col = c("blue","green","red","orange", "grey"), pch = 20)
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
plot(x = mydata$WHOLE, y = mydata$SHUCK, col = c("blue","green","red","orange", "grey")[mydata$CLASS], main = "Shuck weight, in grams|Whole weight, in grams", xlab = "Whole weight",ylab = "Shuck weight", pch = 20)
legend("bottomright", c("A1","A2","A3","A4", "A5"), col = c("blue","green","red","orange", "grey"), pch = 20)
abline(a = 0, b = max(mydata$SHUCK/mydata$WHOLE))

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: There appears to be slightly less variability compared to the plot in (a). Since the shuck weight is a part of the whole weight there is going to be a stronger correlation compared to whole weight and volume. A5 seems to have the lowest shuck weight compared to whole weight followed by A4, A3, and A2. A5 is the oldest age class which can explain why it has the lowest shuck weight compared to whole weight and the trend follows based on the age class. ***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3,3))
hist(mydata$RATIO[mydata$SEX == "F"], col = "red", xlab = "", main = "Female RATIO")
hist(mydata$RATIO[mydata$SEX == "I"], col = "green", xlab = "", main = "Infant RATIO")
hist(mydata$RATIO[mydata$SEX == "M"], col = "blue", xlab = "", main = "Male RATIO")
boxplot(mydata$RATIO[mydata$SEX == "F"], col = "red", xlab = "", main = "Female RATIO")
boxplot(mydata$RATIO[mydata$SEX == "I"], col = "green", xlab = "", main = "Infant RATIO")
boxplot(mydata$RATIO[mydata$SEX == "M"], col = "blue", xlab = "", main = "Male RATIO")
qqnorm(mydata$RATIO[mydata$SEX == "I"], col = "red", main = "Female RATIO")
qqline(mydata$RATIO[mydata$SEX == "I"])
qqnorm(mydata$RATIO[mydata$SEX == "F"], col = "green", main = "Infant RATIO")
qqline(mydata$RATIO[mydata$SEX == "F"])
qqnorm(mydata$RATIO[mydata$SEX == "M"], col = "blue", main = "Male RATIO")
qqline(mydata$RATIO[mydata$SEX == "M"])
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: Based on the displays, the distributions of all three sex classes appear to have a normal distribution. It also appears that all three sex classes are very slightly skewed to the right. This can be explained by the outliers that are evident in the box plot and QQ plot. Overall, all three sex classes have a fairly normal distribution. ***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
inf_out <- boxplot.stats(x = mydata[mydata$SEX == "I", "RATIO"])$out
inf_df <- mydata[mydata$SEX == "I" & mydata$RATIO %in% inf_out, ]
fem_out <- boxplot.stats(x = mydata[mydata$SEX == "F", "RATIO"])$out
fem_df <- mydata[mydata$SEX == "F" & mydata$RATIO %in% fem_out, ]
mal_out <- boxplot.stats(x = mydata[mydata$SEX == "M", "RATIO"])$out
mal_df <- mydata[mydata$SEX == "M" & mydata$RATIO %in% mal_out, ]
kable(inf_df, caption = "Infant")
kable(fem_df, caption = "Female")
kable(mal_df, caption = "Male")
```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: Each sex class appears to have mild outliers but there appears to also be extreme outliers in the female sex class. The total number of ratio outliers was calculated to be 19. There were 8 in the infants, 6 in the females, and 5 in the males.  ***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
par(mfrow = c(2,2))
boxplot(mydata$VOLUME ~ mydata$CLASS, xlab = "", ylab = "VOLUME")
boxplot(mydata$WHOLE ~ mydata$CLASS, xlab = "", ylab = "WHOLE")
plot(x = mydata$RINGS, y = mydata$VOLUME, xlab = "RINGS", ylab = "VOLUME")
plot(x = mydata$RINGS, y = mydata$WHOLE, xlab = "RINGS", ylab = "WHOLE")
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: These variables would be good predictors of age since the greater number of rings an abalone has, the greater its age. This is displayed in the positive correlation of volume and rings and whole and rings scatter plots. As the number of rings increases, the volume and whole increase as well. We can also see this in the box plots. As the age class increases, the volume and whole generally appear to increase along with it.  ***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
mean_vol <- aggregate(VOLUME ~ SEX + CLASS, data = mydata, FUN = mean)
vol_mat <- matrix(data = mean_vol$VOLUME, nrow = 3, byrow = FALSE,
                  dimnames = list(c("Female", "Infant", "Male"), levels(mydata$CLASS)))

mean_shuck <- aggregate(SHUCK ~ SEX + CLASS, data = mydata, FUN = mean)
shuck_mat <- matrix(data = mean_shuck$SHUCK, nrow = 3, byrow = FALSE,
                  dimnames = list(c("Female", "Infant", "Male"), levels(mydata$CLASS)))

mean_ratio <- aggregate(RATIO ~ SEX + CLASS, data = mydata, FUN = mean)
ratio_mat <- matrix(data = mean_ratio$RATIO, nrow = 3, byrow = FALSE,
                  dimnames = list(c("Female", "Infant", "Male"), levels(mydata$CLASS)))

kable(vol_mat, caption = "Volume")
kable(shuck_mat, caption = "Shuck")
kable(ratio_mat, caption = "Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
interaction.plot(x.factor = mydata$CLASS, trace.factor = mydata$SEX, 
                 response = mydata$RATIO, fun = mean, xlab = "Class", ylab = "Ratio", main = "Mean Ratio per Class", col = c("blue", "red", "black"), lty = 1, lwd = 2, trace.label = "SEX")

interaction.plot(x.factor = mydata$CLASS, trace.factor = mydata$SEX, 
                 response = mydata$VOLUME, fun = mean, xlab = "Class", ylab = "Volume", main = "Mean Volume per Class", col = c("blue", "red", "black"), lty = 1, lwd = 2, trace.label = "SEX")

interaction.plot(x.factor = mydata$CLASS, trace.factor = mydata$SEX, 
                 response = mydata$SHUCK, fun = mean, xlab = "Class", ylab = "Shuck", main = "Mean Shuck Weight per Class", col = c("blue", "red", "black"), lty = 1, lwd = 2, trace.label = "SEX")

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (Enter your answer here.)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
par(mfrow = c(2,2))
inf_df5c <- mydata[mydata$SEX == "I" & mydata$RINGS < 16,]
adult_df5c <- mydata[mydata$SEX != "I" & mydata$RINGS < 16,]
boxplot(inf_df5c$VOLUME ~ inf_df5c$RINGS, col = "blue", xlab = "Rings", ylab = "Volume", main = "Infant Volume|Rings", ylim = c(0,1100))
boxplot(adult_df5c$VOLUME ~ adult_df5c$RINGS, col = "red", xlab = "Rings", ylab = "Volume", main = "Adult Volume|Rings", ylim = c(0,1100))
boxplot(inf_df5c$WHOLE ~ inf_df5c$RINGS, col = "blue", xlab = "Rings", ylab = "Whole weight", main = "Infant Whole Weight |Rings", ylim = c(0,400))
boxplot(adult_df5c$WHOLE ~ adult_df5c$RINGS, col = "red", xlab = "Rings", ylab = "Whole weight", main = "Adult Whole Weight|Rings", ylim = c(0,400))
```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: The displays show that abalone volume and whole weight increase as the number of rings increase. It can also be seen that adult abalone growth varies more compared to infant abalone growth. This is evident as the box plots are longer in the adult volume and adult whole weight displays as opposed to the infant volume and whole weight displays.***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: The outliers evident in the data is a plausible statistical reason for the failure of the original study. These outliers created the slight skewness in the results which caused the data to deviate slightly from the normal distribution. This data wouldn't be considered strong enough statistical evidence and would contribute to the failure of the study. In terms of the extent of physical measurements being used for age prediction, the classification of infant needs to be defined as there were multiple infants with numerous rings which should indicate that it is an adult. Also the volume, size, and weight of the abalone can be used to determine the age of the abalone to an extent. The environment of the abalone needs to be considered as well since that is another factor that directly contributes to the measurements of those variables. ***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: I would need to first determine the size of the sample, normality of the sample, standard deviation of the sample, the mean of the sample, and the number of outliers in the sample.***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: The difficulty in analyzing data derived from observational studies is determining if the data is non-bias and accurate. This needs to be determined because observational data can a lot of times be not collected properly or have some sort of errors in the data sample. Causality can also not be determined even though there is a strong correlation between the variables. The relationship between the variables is unknown which means other underlying factors could be contributing to the correlation between them. Hidden correlations can be learned from these studies between unexpected variables which would prompt you to analyze and research them further to better understand the relationship between them. ***
